---
title: "Lab2HDS"
author: "Benjamin Wilson"
date: "2026-09-15"
output: html_document
---

## R Markdown setup

Came with the file.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading the data

This section identifies the url for the RNA-seq data and downloads the file to the 
working directory. This differs from the Source.md by downloading the file before 
creating the fpkm object, for my computer was having issues creating the variable
before. 


```{r url}
url <- "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE52778&format=file&file=GSE52778_All_Sample_FPKM_Matrix.txt.gz"

download.file(url, "GSE52778_FPKM.txt.gz", mode = "wb")
```

## Creating the FPKM table

With the dataset downloaded, the fpkm object can be created. As stated before, the
code before did not work for me, so I had to download the file first. I asked 
GPT-4o-mini for assistance in creating the new object. This new table takes the
text file, specifies there are headers, creates separators, and makes sure the data
is not treated as factors. Then, two functions check for quality control with 
dimensions and the first few rows of the table.


```{r}
fpkm <- read.table(
  "GSE52778_FPKM.txt.gz",
  header = TRUE,
  sep = "",
  stringsAsFactors = FALSE
)

dim(fpkm)
head(fpkm)
```

## Finding the top 10 genes with the highest Dex FPKM

I wanted to find the 10 genes with the highest Dex FPKM values. The code below 
orders the fpkm data frame by the Dex_FPKM column in descending order and then 
selects the top 10 rows. It then ties this value to the gene_short_name column.


```{r}
top_genes <- fpkm[order(fpkm$Dex_FPKM, decreasing = TRUE), ]
top_genes <- head(top_genes, 10)

top_genes[, c("gene_short_name", "Dex_FPKM")]
```

## Creating a bar plot of the top 10 genes

I used my familiarity with the ggplot2 package to create a bar plot of the top 10 
genes with the highest Dex FPKM values. The code below specifies the data frame, 
the x and y aesthetics, and uses geom_col to create the bars. It also adds labels 
and a minimal theme for better visualization.

```{r}
library(ggplot2)

ggplot(top_genes, aes(x = gene_short_name, y = Dex_FPKM)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Top 10 Most Highly Expressed Genes in Dex",
    x = "Gene",
    y = "FPKM"
  ) +
  theme_minimal()
```
